---
title: "EDS 223 Assignment 1"
subtitle: "Map making practice"
author: "Monique Hernandez"
date: 2026-10-3
execute:
  eval: true
  warning: false
  message: false
format: pdf
---

# Air Quality and Cancer risk in Oahu, Hawaii via EJScreen Tool

The two maps were created to focus on the environmental injustice for those living on the island of Oahu in Hawaii that may face a higher risk of cancer from toxic chemicals in the air due. The first map displays the percentage of high risk of cancer, as listed by the EPA's Toxics Release Inventory (TRI) Program, which is mostly concentrated at the bottom of Oahu. The second map focuses on the concentration chemicals modeled by the toxicity weight.

Map 1 determines the toxic chemical weight, with the EJscreen having raw values range from 0 to 4,512,172.1, with the highest at 28,117.14. Map 2 shows the percentile of high risk areas in danger of cancer from toxic chemicals, which was calculated by persons per million over a 70-year lifetime with the max being 50 of a raw value.

The burden of those living near the areas with high risk of cancer may impact those who face poverty, which could be further mapped to show those living in the vicinity with the highest percetange faciing high risk of cancer. Lastly, it's important to note that map 1 and map 2 have a strong correlation with the location of two major infrastructures on the island that pollute the island. These two are the Daniel K. Inouye International Airport and Par Hawaii Refining,an energy plant, that had a recent settlement in 2022 due to Clean Air Violations with the EPA. These two areas are also major tourist attractions, with a strip of hotels and multiple shopping areas increasing the amount of anthropogenic pollution emitted on the island.

## Load packages
```{r}
library(tidyverse) # For R
library(sf) # Read vector 
library(here) # Locate files
library(stars)
library(tmap) # For static and interactive maps
library(janitor) # Clean data
``` 


## Read EJScreen and Data
```{r}
# read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read(here::here("data",
"ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"),
quiet = TRUE) 

# filter to a state you are interested in
hawaii <- ejscreen |> 
  dplyr::filter(ST_ABBREV == "HI") 

# filter to a county you are interested in
honolulu <- ejscreen |> 
  dplyr::filter(CNTY_NAME %in% c("Honolulu County"))

# find the average values for all variables within counties
hawaii_counties <- aggregate(hawaii,
by = list(hawaii$CNTY_NAME), FUN = mean)

```

## Data Cleaning
```{r}
# Convert all columns to lower case
hawaii <- hawaii |>
  clean_names()

honolulu <- honolulu |>
  clean_names()

hawaii_counties <- hawaii_counties |>
  clean_names()

```


## Map 1: Percentage of Lifetime Risk of Cancer from Toxic Air
```{r}
# Select primary spatial object as county of choice
m1 <- tm_shape(honolulu,
bbox = c(-158.3042, 21.2367, -157.6251, 21.7293)) +
  tm_graticules() +
  tm_polygons(fill = "cancer", # Map percentile of air cancer
    fill.legend = tm_legend(title = "Percentage %", # Create legend
    bg.color = "skyblue",
    width = 8),
    fill.scale = tm_scale_intervals(
      values = c("green", "yellow", "red"),
      value.na = "skyblue")) +
  tm_compass( # Customize compass
    type = "rose",
    size = 3,
    position = c(
      "right", "top")) +
  tm_scalebar(position = c("left", "bottom")) + 
  tm_title(
      text = "Percentage of Lifetime Cancer Risk from Toxic Air in Oahu",
      size = 1) + # Add title to map
  tm_layout(scale = 1.2, 
    asp = 0,    # Customize backgrounds
    style= "beaver",
    bg.color = "skyblue",
    outer.bg.color = "lightyellow") +
  tm_credits("Data source: OAQPS (2019)",
    fontface = "italic",
    position = c("left", "top"), size = .5) +
  tm_logo("image/epa.png", height=1,
    position = c("right", "bottom"))
m1
```

## Map 2: Toxic Air Chemicals 

```{r}
m2 <- tm_shape(honolulu,
bbox = c(-158.3042, 21.2367, -157.6251, 21.7293)) +
  tm_graticules() +
  tm_polygons(fill = "rsei_air",
    fill.legend = tm_legend(title = "Concentration",
    bg.color = "skyblue",
    width = 8),
    fill.scale = tm_scale_intervals(
      values = c("green", "yellow", "red"),
      value.na = "skyblue")) +
  tm_compass(
    type = "rose",
    size = 2,
    position = c("right", "top")) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_title(text = "Concentration of Toxic Chemicals in Oahu",
    size  = 1) + 
  tm_layout(scale= 1.2,
    asp = 0,
    style = "beaver",
    bg.color = "skyblue",
    outer.bg.color = "lightyellow") +
  tm_credits("Data source: EjScreen (2019)",
    fontface = "italic", position = c("left", "top")) +
  tm_logo("image/epa.png", height=1,
    position = c("right", "bottom"))
m2
```
